P4IRS: An Intermediate Representation and Compiler for Parallel and Performance-Portable Particle Simulations

### 2025-05-02 sage_latex_template/manuscript.tex
Add text about loop splitting
Signed-off-by: Rafael Ravedutti <rafael.r.ravedutti@fau.de>

% Acronyms
\newacro{AoS}{array of structures}
\newacro{AoSoA}{array of structures of arrays}
\newacro{CPI}{cycles per instruction}
\newacro{DEM}{discrete element method}
\newacro{EAM}{embedded atom method}
\newacro{FCC}{face-centered cubic}
\newacro{FN}{full neighbor-lists}
\newacro{HN}{half neighbor-lists}
\newacro{ILP}{instruction-level parallelism}
\newacro{IR}{intermediate representation}
\newacro{AST}{abstract syntax tree}
\newacro{ISA}{instruction set architecture}
\newacro{LJ}{Lennard-Jones}
\newacro{MD}{Molecular dynamics}
\newacro{MPI}{message passing interface}
\newacro{SIMD}{single instruction, multiple data}
\newacro{SoA}{structure of arrays}
\newacro{PBC}{periodic boundary conditions}
\newacro{HPM}{hardware performance monitoring}

@@

\subsection{Intermediate Representation}

@@

An important aspect of DSLs is to be able to use domain knowledge for their compiler transformations, thus enabling code optimizations that may not be detected by general purpose compilers.
In order to achieve this, P4IRS provides its own IR which can represent elements that are specific to particle computations.
These particle-specific elements in the AST can then be used in P4IRS compilation passes, and at some point during the compilation are lowered, resulting in the low level AST that corresponds to the code implementation that matches the strategies given by the user.}

@@

\subsection{Runtime Library}

@@

To provide a simpler way to achieve vectorized code for such DEM kernels, we partition particles by shape in our acceleration data structures (Linked Cells and Verlet Lists), which allows distinguishing them during the force computation.
After, we use code generation to apply a loop-splitting in the generated code for every pair of shapes, these resulting loops should be then easier to be vectorized since no branches are needed to calculate different contact elements for the particles being computed.
Algorithm \ref{alg:no_branching} displays the same kernel as Algorithm \ref{alg:branching} without the branching issue.
Note that the outer-most loop can also be split into two parts as well, which leads to two different kernels for each pair of shapes.
For GPUs, this can be beneficial since different threads will be used for different pair of shapes (parallelism is done at the level of the outer-most loop) and no threads will be disabled due to traversing a different code path.

@@

>>>>

% Acronyms
\newacro{AoS}{array of structures}
\newacro{AoSoA}{array of structures of arrays}
\newacro{CPI}{cycles per instruction}
\newacro{DEM}{discrete element method}
\newacro{EAM}{embedded atom method}
\newacro{FCC}{face-centered cubic}
\newacro{FN}{full neighbor-lists}
\newacro{HN}{half neighbor-lists}
\newacro{ILP}{instruction-level parallelism}
\newacro{IR}{intermediate representation}
\newacro{DSL}{domain-specific language}
\newacro{AST}{abstract syntax tree}
\newacro{ISA}{instruction set architecture}
\newacro{LJ}{Lennard-Jones}
\newacro{MD}{Molecular dynamics}
\newacro{MPI}{message passing interface}
\newacro{SIMD}{single instruction, multiple data}
\newacro{SoA}{structure of arrays}
\newacro{PBC}{periodic boundary conditions}
\newacro{HPM}{hardware performance monitoring}

@@

\subsection{\RMchange{Intermediate Representation}}

@@

An important aspect of a \ac{DSL} is to be able to use domain knowledge for their compiler transformations, thus enabling code optimizations that may not be detected by general purpose compilers.
In order to achieve this, P4IRS provides its own IR which can represent elements that are specific to particle computations.
These particle-specific elements in the AST can then be used in P4IRS compilation passes, and at some point during the compilation are lowered, resulting in the low level AST that corresponds to the code implementation that matches the strategies given by the user.}

@@

\subsection{Runtime \RMchange{Library}}

@@

To provide a simpler way to achieve vectorized code for such DEM kernels, \RMchange{it is possible to} partition particles by shape in the acceleration data structures (Linked Cells and Verlet Lists), which allows distinguishing them during the force computation.
After, \RMchange{multiple loops can be used} in the code for handling different pair of shapes, \RMchange{and} \RMchange{vectorizing} these resulting loops should be easier since no branches are needed to calculate different contact elements for the particles being computed.
Algorithm \ref{alg:no_branching} displays the same kernel as Algorithm \ref{alg:branching} without the branching issue.
Note that the outer-most loop can also be split into two parts as well, which leads to two different kernels for each pair of shapes.
For GPUs, this can be beneficial since different threads will be used for different pair of shapes (parallelism is done at the level of the outer-most loop) and no threads will be disabled due to traversing a different code path.

@@

\RMchange{P4IRS is able to generate this algorithmic optimization automatically for the shapes involved in the simulation.
This is possible by using the domain-knowledge provided in the AST, where the \emph{ParticleInteraction} symbol can be lowered to several loops that handle different pairs of shapes in the simulation.
Note that P4IRS does not require complex and sophisticated analyses to identify these opportunities for splitting the interaction loop because its \ac{IR} already exposes the domain-knowledge necessary for applying this strategy.
As discussed, this is an important benefit of using a \ac{DSL} because the \ac{IR}s from general-purpose compilers need to work at a lower level in order to be able to represent any generic code.}
